Figures

### 2016-03-16 doc/paper/results.Rmd
starting methods

DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants. We asked whether the decreased accuracy observed in our data set was due simply topeculiarities specific to the DeepSNV algorithm. We therefore analyzed our $10^5$ input populations using Lofreq, another alogorthim that is commonly used in viral next generation sequencing studies. Lofreq has been reported to have perfect specificity in the past. Under our conditions Lofreq had marginally reduced sensitivity compared DeepSNV when applied to the variant frequencies \geq 1.0% but marginally increased sensitivity when applied to variant frequencies \l 1.0% (`r Lofreq`). The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering). This increased specificity is most likely due to the fast that  the LoFreq algorithm takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants fount only at one end of the reads. However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`) with over 40 false positives per sample. It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague all variant callers applied to patient-derived viral samples.

>>>>

DeepSNV is one of many variant callers that use a combination of empiric and statistical approaches to model error rates and separate true from false variants. We asked whether the decreased accuracy observed in our data set was due simply topeculiarities specific to the DeepSNV algorithm. We therefore analyzed our $10^5$ input populations using Lofreq, another alogorthim that is commonly used in viral next generation sequencing studies. Lofreq has been reported to have perfect specificity in the past. Under our conditions Lofreq had marginally reduced sensitivity compared DeepSNV when applied to the variant frequencies $\geq$ 1.0% but marginally increased sensitivity when applied to variant frequencies $\l$ 1.0% (`r Lofreq`). The specificity of LoFreq was comparable to what we observed with DeepSNV in our high-input cell culture-derived populations (`r Miseq`), and better than DeepSNV in our initial implementation (`r Hiseq` prior to Phred, MapQ, and read position filtering). This increased specificity is most likely due to the fast that  the LoFreq algorithm takes MapQ and Phred scores into account when calling variants and has a stringent strand bias filter that removes many of the variants fount only at one end of the reads. However even with these additional characteristics, the specificity of LoFreq was lower than our improved DeepSNV pipeline (`r Qual`) with over 40 false positives per sample. It appears that higher than expected false positive rates are not specific to DeepSNV and most likely plague all variant callers applied to patient-derived viral samples.

### 2016-03-16 scripts/figures.Rmd
starting methods

```{r,Miseq_read.cut,eval=F}
### With read position cut off 63 and 188
read.df<-subset(sum.df,Read_pos>63 & Read_pos<188)
m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars,">")

@@

sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output

# Getting genome lenght and segment information from the wsn33 fasta file
reference.fasta<-"../data/reference/wsn33_wt_plasmid.fa" 
segments <- fasta.seqlengths(reference.fasta)
regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=12, stop=segments-13, row.names=NULL) # the univeral primers are 12 and 13 bp long
regions.bed<-mutate(regions.bed,chr=as.character(chr))

@@

#length(wt_mut) How many wt muts do we find.
sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT samples

@@

possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position

hiseq_vars<-possible_vars
hiseq_kb<-sum(regions.bed$length)/1000

@@

@@

direction<-">"
lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")

@@

ggdraw() +
  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
  draw_plot(tbl, 0.025, 0, 0.95, .48) +
  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
direction<-"<"
```

@@

euc<-function(x,ref){
  x$Id="test"
  ref$Id="ref"
  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),subset(ref,select=c(mutation,freq.var,Id)))
  tab<-as.matrix(subset(dcast(all,mutation ~ Id, value.var = "freq.var"),select=-c(mutation)))
  tab[is.na(tab)]<-0
  dist(t(tab),"manhattan")[1]
}
get_data<-function(x,setting){
  if(length(unique(x$freq.var))==1){
    exp.freq=unique(x$freq.var) # this is the perfect condition where there is only 1 frequency and it is all the same
  }else{
  exp.freq=unique(x$exp.freq)
  }
  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),"1/D"=D(x),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3))
  rownames(out)<-setting
  return(t(out))
}
tabulate<-function(input,freq){
get_data(subset(perfect,freq.var==freq),"Perfect")->perf
get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV")->deepSNV
get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01),"DeepSNV Final")->deepSNV.f
knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f))
}
# tabulate.4<-function(input,freq){
# get_data(subset(perfect,freq.var==freq),"Perfect")->perf
# get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
# get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
# get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
# get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
# 
# knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
# }

>>>>

```{r,Miseq_read.cut,eval=F}
### With read position cut off 63 and 188
read.df<-subset(sum.df,Read_pos>62 & Read_pos<188)
m.roc.df<-miseq.roc(read.df,possible_tp,possible_vars,">")

@@

sum.df<-read.csv("../data/process/2015-6-23/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#") # reading in the variants from the pipeline output
mfe.df<-read.csv("../data/reference/mfe.csv") 
# Getting genome lenght and segment information from the wsn33 fasta file
reference.fasta<-"../data/reference/wsn33_wt_plasmid.fa" 
segments <- fasta.seqlengths(reference.fasta)
regions.bed <- data.frame(chr = gsub("[ ].*","", names(segments)), start=12, stop=segments-13, row.names=NULL) # the univeral primers are 12 and 13 bp long
regions.bed<-mutate(regions.bed,chr=as.character(chr))

@@

#length(wt_mut) How many wt muts do we find.
sum.df$category[sum.df$mutation %in% wt_mut]<-"wt"
sum.df<-subset(sum.df,!(grepl("WT",Id))) # remove the WT sample

sum.df$category[sum.df$mutation %in% mfe.df$mutation]<-"mfe"

@@

total_possible_vars<-sum(regions.bed$length)*3 # three possible variants/ position

possible_vars<-total_possible_vars-length(wt_mut)-dim(mfe.df)[1] # removing the mfe and wt_mut sites
hiseq_vars<-possible_vars
hiseq_kb<-sum(regions.bed$length)/1000

@@

lofreq.df$category[lofreq.df$mutation %in% wt_mut]<-"wt"
lofreq.df$categroy[lofreq.df$mutation %in% mfe.df$mutation]<-"mfe"

@@

lofreq.roc<-hiseq.roc(lofreq.df,possible_tp,possible_vars,"<")

@@

ggdraw() +
  draw_plot(lofreq.roc.p, 0, 0.5, 0.9, 0.45) + 
  draw_plot(tbl, 0.025, 0, 0.95, .48) +
  draw_plot_label(c("A", "B"), c(0, 0), c(1, 0.5), size = 15)
```

@@

euc<-function(x,ref){
  x$Id="test"
  ref$Id="ref"
  all<-rbind(subset(x,select=c(mutation,freq.var,Id)),subset(ref,select=c(mutation,freq.var,Id)))
  tab<-as.matrix(subset(dcast(all,mutation ~ Id, value.var = "freq.var"),select=-c(mutation)))
  tab[is.na(tab)]<-0
  dist(t(tab),"manhattan")[1]
}
get_data<-function(x,setting){
  if(length(unique(x$freq.var))==1){
    exp.freq=unique(x$freq.var) # this is the perfect condition where there is only 1 frequency and it is all the same
  }else{
  exp.freq=unique(x$exp.freq)
  }
  out<-data.frame(Richness=round(rich(x),digits = 0), Entropy=format(H(x),scientific = T,digits = 3),Manhattan=round(euc(x,subset(perfect,freq.var==exp.freq)),digits=3)) #"1/D"=D(x),
  rownames(out)<-setting
  return(t(out))
}
tabulate<-function(input,freq){
get_data(subset(perfect,freq.var==freq),"Expected")->perf
get_data(subset(sum.df,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F)),"DeepSNV")->deepSNV
get_data(subset(lofreq.df,gc==input & exp.freq==freq & category %in% c(T,F)),"Lofreq")->lo
get_data(subset(cut.read.df.all,gc==input & exp.freq==freq & p.val<0.01 & category %in% c(T,F) ),"DeepSNV Final")->deepSNV.f
knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f))
}
# tabulate.4<-function(input,freq){
# get_data(subset(perfect,freq.var==freq),"Perfect")->perf
# get_data(subset(sum.df,gc==input & exp.freq==freq),"DeepSNV")->deepSNV
# get_data(subset(lofreq.df,gc==input & exp.freq==freq),"Lofreq")->lo
# get_data(subset(cut.read.sig.df,gc==input & exp.freq==freq),"DeepSNV Final")->deepSNV.f
# get_data(subset(dups,exp.freq==freq),"DeepSNV duplicates")->deepSNV.d
# 
# knitr::kable(cbind(perf,deepSNV,lo,deepSNV.f,deepSNV.d),digits = 2)
# }
